Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-04-04 paper/main.tex
equivariance
cleared out, rewrote intro to math section

\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}

\IEEEPARstart{T}his paper uses methods from algebraic topology and category theory to develop a model of the transformation from data to graphical representation. This model provides a language to specify how data is structured and how this structure is carried through in the visualization, and serves as the basis for a functional approach to implementing visualization library components. Topology allows us to describe the structure of the data and graphics in a generalizable, scalable, and trackable way. Category theory provides a framework for separating the transformations implemented by visualization libraries from the various stages of visualization and therefore can be used to describe the constraints imposed on the library components \cite{wielsManagementEvolvingSpecifications1998,goguenCategoricalManifesto1991}. Well constrained modular components are inherently functional\cite{hughesWhyFunctionalProgramming1989}, and a functional framework yields a library implementation that is likely to be shorter, clearer, and more suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}. Using this functional approach, this paper contributes a practical framework for decoupling data processing from visualization generation in a way that allows for modular visualization components that are applicable to a variety of data sets in different formats. \note{is it OK that this is something reviewer 4 wrote}

\section{Related Work}
This work aims to develop a model for describing visualization transformations that can serve as guidance for how to architecture a general purpose visualization library. We define a general purpose visualization library as one that provides non domain specific building block components\cite{wongsuphasawatNavigatingWideWorld2021} for building visualizations, for example functions for converting data to color or encoding data as dots. We propose that visualization library components should preserve continuity and equivariance, which we define as

\begin{LaTeXdescription}
  \item [continuity] how elements in a dataset are organized, e.g. discrete rows in a table, networked nodes, pixels in an image, points on a line
  \item [equivariance] data transformations that have an equivalent effect on the graphical representation, e.g. rotating a matrix has a matching rotation of the image, translating the points on a line has a matching visual shift in the line plot
\end{LaTeXdescription}

In \autoref{sec:related-work:continuity} we discuss how data continuity influences visualization library architecture and in \autoref{sec:related-work:equivariance} we summarize how visualization researchers have approached equivariance.

\subsection{Continuity}
\label{sec:related-work:continuity}
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{k_different_types.png}
  \caption{Continuity is how elements in a data set are connected to each other, which is distinct from how the data is structured. The rows in (a) are discrete, therefore they have discrete continuity as illustrated by the discrete dots. The gaussian in (b) is a 1D continuous function, therefore the continuity of the elements of the gaussian can be represented as a line on an interval (0,1). In (c), every element of the globe is connected to its nearest neighbors, which yields a 2D continuous continuity as illustrated by the square.}
  \label{fig:related-work:continuity}
\end{figure}
Continuity describes elements in a data set are organized; this concept is termed topological properties by Wilkinson\cite{wilkinsonGrammarGraphics2005}. Wilkinson provides the examples of values that are isolated from each other, and therefore discrete, and values lying on a continuum or in a compact region. For example, in \autoref{fig:related-work:continuity}, each station record in the table is independent of the others; therefore, the continuity of the table is discrete. The data provided by the gaussian are points sampled along the curve, therefore the continuity of the points on the line is 1D continuous. Every point on the globe is connected to its 6 nearest cardinal neighboring points (NW, N, NE, E, SE, S, SW, W). We propose that a robust model of continuity provides a way to develop library components that can work on the data in small pieces in a manner where the overall continuity of the data is preserved in the visual transformation.

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{whycontinuity.png}
  \caption{Continuity is implicit in choice of visualization rather than explicitly in choice of data container. The line plots in (b) are generated by a 2D table (a). Structurally this table can be identical to the 2D matrix (a) that generates the image in (c).}
  \label{fig:related-work:visual-algorithm}
\end{figure}

 The preservation of continuity can be made explicit, as in the transformation of table to parallel coordinates in Ruchikachorn and Mueller \cite{ruchikachornLearningVisualizationsAnalogy2015}, but is often expressed implicitly in the choice of visual algorithm (visualization type), as explored in taxonomies by Tory and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004} and Chi\cite{chiTaxonomyVisualizationTechniques2000}.

For example, in \autoref{fig:related-work:visual-algorithm} the same table can be interpreted as a set of 1D continuous curves when visualized as a collection of line plots or as a 2D surface when visualized as an image.  This means that often there is no way to express data continuity independent of visualization type, meaning most visualization libraries will allow, for example, visualizing discrete data as a line plot or an image.

\subsection{Equivariance}
\label{sec:related-work:equivariance}
When introducing the retinal variables, Bertin informally specifies that continuity is preserved in the mark and defines equivariance constraints in terms of data and visual variables being selective, associative, ordered, or quantitative\cite{bertinSemiologyGraphicsDiagrams2011a}. In the \textit{A Presentation Tool}(APT) model, Mackinlay embeds the continuity constraint in the choice of visualization type and generalizes the equivariance constraint to preserving a binary operator from one domain to another. The algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}, proposed by Kindlmann and Scheidegger, restricts equivariance to invertible transformations.

\begin{figure}[h!]
  \label{fig:related-work:equivariance}
  \includegraphics[width=\columnwidth]{equiv.png}
  \caption{Equivariance is that a transformation on the data has a corresponding transformation in the graphical representation. For example, in this figure the data is scaled by a factor 10. Equivalently the line plot is scaled by factor of 10, resulting in a shrunken line plot. Either a transformation on the data side can induce a transformation on the visual side, or a transformation on the visual side indicates that there is also a transformation on the data side. }
\end{figure}

\subsection{Visualization Library Architecture}
General purpose visualization libraries-such as Matplotlib\cite{hunterMatplotlib2DGraphics2007}, Vtk\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012}, and D3 \cite{bostockDataDrivenDocuments2011}-carry distinct data models as part of the implementation of each visual algorithm. The lack of unified data model means that each plot in a linked\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991} visualization is treated as independent, as are the transforms converting each field in the data to a visual equivalent.

Domain specific libraries can often guarantee consistency because they have a single model of the data in their software design, as discussed in Heer and Agarwal \cite{HeerSoftware2006}'s survey of visualization software design patterns. For example, the relational database is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, Vega\cite{satyanarayanDeclarativeInteractionDesign2014} and Altair\cite{vanderplasAltairInteractiveStatistical2018}. Images underpin scientific visualization tools such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and the digital humanities oriented ImagePlot\cite{studiesCulturevisImageplot2021} macro; the need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}.

\section{Algebraic Topology \& Category Theory}
\note{potentially move this to appendix, push Butler \& Vickers up to RW}

Category theory can be used to express software specifications because it provides a method of expressing structure and how that structure can be composed \cite{wielsManagementEvolvingSpecifications1998} algebraic topology provides a rich language for expressing data continuity. In this section, we introduce concepts from algebraic topology and category theory that we will use in \autoref{sec:artist:category} to develop a formal model of specifying visualization component.

\subsection{Fiber Bundles}
\label{sec:related-work:fiber-bundles}
\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{fiberbundle.png}
  \caption{A fiber bundle is mathematical construct that allows us to express the relationship between data and continuity. The \textcolor{total}{total} space \dtotal is the topological space in which the data is embedded. The \textcolor{fiber}{fiber} space \dfiber\ is embedded in \dtotal\ and is the set of all possible values that any
  \note{add big rectangle E}}
  \label{fig:related-work:fiber-bundle}
\end{figure}

The model described in this work provides a model for expressing data sets with different topological properties.

\subsection{Fiber Bundles}
We obtain this generality by using a mathematical structure called a fiber bundles as the basis of our abstraction, as proposed by Butler, Bryson, and Pendley\cite{butlerVisualizationModelBased1989,
butlerVectorBundleClassesForm1992}. As described by them, a fiber bundle is a formal model of the mapping between data points and the underlying topological space they lie in. For example, nodes on a network and the graph of the network, an image and the underlying grid at which the image is sampled, or table columns and the table index.

Formally, a fiber bundle is a mathematical structure $(\dtotalc, \dbasec, \pi, \dfiberc)$
\begin{equation}
  \label{eq:fiber_bundle}
  \begin{tikzcd}
      \dfiberc \arrow[r, hook] & \dtotalc \arrow[r, "\pi"] & \dbasec
  \end{tikzcd}
\end{equation}

\begin{definition}
The \textcolor{base}{base space} \dbasec\ is a topological space, which means it is a set $\dbase$ with a collection of open sets \cite{TopologyVsTopology} surrounding each element in the set. Open sets are a collection of subsets $\{\opensetc\}$ in a set \dbase, including the empty set, the whole set \dbase, and every union and intersection of its member subsets. For each point $\dbasepoint \in \dbase$, there is a collection of subsets $\{\openset\} \subset \dbase$ such that $\dbasepoint \in \openset \subseteq \dbase$ \cite{spanier1989algebraic, bradleyTopologyCategoricalApproach2020}  
\end{definition}
\note{intuition about base space}

\begin{definition} The \textcolor{fiber}{fiber space} \dfiberc\ is a topological space such that for every $\dbasepoint \in \dbase$, the fiber over that point is isomorphic to the preimage of that point $\dfiber_{\dbasepoint} \cong \pi^{-1}(\{\dbasepoint\})$. All fibers $\dfiber_{\dbasepoint}$ are homeomorphic to each other \cite{spanier1989algebraic, hatcherAlgebraicTopology2002}
\end{definition}

The fibers can be thought of as encapsulating data types according to Spivak, who proposes that databases can be represented as fiber bundles \cite{spivakSIMPLICIALDATABASES,spivakDatabasesAreCategories2010}. Spivak formulates the fiber as a (field name, data domain) simple schema where field names and field types can be formally mapped to the set of values associated with the field type, for example $\mathbb{R}$ for a \texttt{float} column named \textit{temperature}.

\begin{definition} The \textcolor{total}{total space} \dtotalc\ is the space reachable via the projection map $\pi$. The total space is always locally trivial, meaning $\dtotal = \dbase \times \dfiber$ over any open neighborhood $\openset_{\dbasepoint}$. 
\end{definition}

The fiber bundle is trivial when $\dtotal = \dbase \times \dfiber$ is true globally; a non-trivial bundle can be thought of as the twisted fiber product $\dtotal=\dbase times_{T} \dfiber$\cite{hatcherAlgebraicTopology2002,nlab:twisting_function}. The twisting refers to the fiber \dfiber\ not aligning in the same direction over all $\openset \subseteq \dbase$.

 There is a map from a point on the base space \dbase\ which returns values in the fiber space \dfiber. This map is called a section $\dsection: \dbase \rightarrow \dtotal$ of the fiber bundle

\begin{equation}
  \label{eq:related-work:continuity:fiber-bundle}
  \begin{tikzcd}[row sep=large]
      \dfiberc \arrow[r, hook] & \dtotalc \arrow[d, "\pi"'] \\
                        & \dbasec \arrow[u, "\dsectionc \in \cgamma{\dbasec}{\dtotalc}"', bend right, pos=.5]
  \end{tikzcd}
\end{equation}

and the set of all sections over \dbase\ is denoted $\Gamma(\dbase, \dtotal)$ . Sections can also be defined locally, meaning over an open set $\openset \subseteq \dtotal$ and the set of all sections over an arbitrary open set is $\Gamma(\openset, \dtotal)$. 
\begin{equation}
  \label{eq:related-work:continuity:sections}
  \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\openset} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;\forall \dbasepointc \in \opensetc \big\}
\end{equation}
Butler et al proposed that \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992} sections could be used to model data sets since they are maps from a point on the continuity to a point in the fiber, and therefore they encapsulate both. The set of sections is the space of all data sets with the same continuity and fields.

\subsection{Category Theory}
\label{sec:related-work:equivariance:category}
In this work, we propose that equivariance constraints can be expressed using category theory. Vickers et. al \cite{vickersUnderstandingVisualizationFormal2013} provide a brief introduction to category theory for visualization practitioners, but their work focuses on reasoning about visualization design, while this paper aims to provide guidance on designing visualization library components. Briefly we introduce concepts in category theory that we use to describe our model in \autoref{sec:artist:category} and \autoref{sec:artist:construction}.

\begin{definition} A \textcolor{source}{category} $\textcolor{source}{\mathcal{C}}$ consists of a collection of objects $c$ with identity $id_c: C\rightarrow C$ and the set of morphisms between every two objects $f:C_1 \rightarrow C_2$, \cite{fongInvitationAppliedCategory2019,maclaneCategoriesWorkingMathematician2013}. 
\end{definition}
The set of morphisms between objects is termed the hom set $hom_{\mathcal{C}}(C_1, C_2)$. The morphisms on the category  compose
\begin{equation*}
  \begin{tikzcd}
    \textcolor{source}{C_1} \arrow[r, "f", color=source] \arrow[rd, "g\circ f"', color=source] \arrow["id_{C_1}"', loop, distance=2em, in=215, out=145, color=source] & \textcolor{source}{C_2} \arrow[d, "g", color=source] \arrow["id_{C_2}"', loop, distance=2em, in=125, out=55, color=source] \\
  & \textcolor{source}{C_3} \arrow["id_{C_3}"', loop, distance=2em, in=305, out=235, color=source]              
  \end{tikzcd}
\end{equation*}
 and are constructed such that the following axioms hold \cite{riehlCategoryTheoryContext}:
 \begin{LaTeXdescription}
   \item[associativity] if $f: C_1 \rightarrow C_2$, $g: C_2 \rightarrow C_3$ and $h: C_3 \rightarrow C_4$ then $h\circ (g \circ f) = (h \circ g) \circ f$
   \item[identity] for every $f: C_1 \rightarrow C_2$ there exists identity morphisms $f \circ id_{C_1} = f = id_{C_2} \circ f$
 \end{LaTeXdescription}

 \begin{definition} An opposite category $\mathcal{C}^{op}$ is a category with all the same objects of category $\mathcal{C}$ and morphisms that are reversed. For example $f:C_1 \rightarrow C_2$ in $\mathcal{C}$ is $f:C_2 \rightarrow C_1$ in $\mathcal{C}^{op}$.   
 \end{definition}

 \begin{definition} A functor is a morphism between objects of any two  categories $\textcolor{sheaf}{\bm{F}}: \textcolor{source}{\mathcal{C}} \rightarrow \textcolor{target}{\mathcal{D}}$. A functor has the properties of identity and composition \cite{riehlCategoryTheoryContext}\end{definition}

A functor preserves the structure of the category, namely morphisms and the source and target objects of those morphisms, composition of morphisms, and identity morphisms \cite{riehlCategoryTheoryContext}. For example, the functor $\textcolor{functor}{F}:\mathcal{C} \textcolor{functor}{\rightarrow} \mathcal{D}$ maps an object in $\textcolor{source}{\mathcal{C}}$ into an object in $\textcolor{target}{\mathcal{D}}$
\begin{equation}
\label{eq:related-work:equivariance:functor_diagram}
\begin{tikzcd}
  \textcolor{source}{c} \arrow[d, "f"', shift right=0, color=source] \arrow[r, "F", color=functor] & \textcolor{target}{F(c)} \arrow[d, "F(f)", shift right=0, color=target] \\
  \textcolor{source}{c^{\prime}} \arrow[r, "F"', color=functor]                       & \textcolor{target}{F(c^{\prime})}                      
  \end{tikzcd}
\end{equation}
 such that for every morphism $f: c \rightarrow c^{\prime} \in Hom(\mathcal{C})$ there is a compatible morphism $F(f): F(c) \rightarrow F(c^{\prime}) \in Hom(\mathcal{D})$. Functors $\bm{F}: \mathcal{C}^{op} \rightarrow \mathcal{D}$ are called contravariant functors.

 \begin{definition} A \textcolor{nattran}{natural transformation} is a morphism of functors $\textcolor{nattran}{\alpha}: \textcolor{functor}{F} \textcolor{nattran}{\rightarrow} \textcolor{functor}{G}$ where $F, G$ are functors from $\mathcal{C}$ to $\mathcal{D}$. \cite{riehlCategoryTheoryContext, bradleyWhatNaturalTransformation}
 \end{definition}

\begin{equation}
  \label{eq:related-work:equivariance:natural-transform}
\begin{tikzcd}
  & {} \arrow[dd, "\alpha", Rightarrow, shorten=1.5em, color=nattran] &             \\
\textcolor{source}{\mathcal{C}} \arrow[rr, "F", bend left, color=functor] \arrow[rr, "G"', bend right, color=functor] &                                     & \textcolor{target}{\mathcal{D}} \\
  & {}                                  &            
\end{tikzcd}
\end{equation}
The natural transformation composes with a functor such that the output is a compatible functor between the same categories. This means that for every morphism $f: c \rightarrow c^{\prime}$ between objects in $\mathcal{C}$, there is a commutating naturality square for the corresponding objects in $\mathcal{D}$ \cite{milewskiCategoryTheoryProgrammers}. 
\begin{equation}
  \label{eq:related-work:equivariance:natural_transform_commute}
  \begin{tikzcd}[column sep=huge]
    \textcolor{target}{F(c)} \arrow[dd, "F(f)"', color=target] \arrow[rr, "\alpha_{c}", bend left, color=nattran] & \textcolor{source}{c} \arrow[dd, "f" description, color=source] \arrow[r, "G"', color=functor] \arrow[l, "F", color=functor] & \textcolor{target}{G(c)} \arrow[dd, "G(f)", color=target] \\
    {} \arrow[rr, dotted, color=nattran] & & {}                      \\
    \textcolor{target}{F(c^{\prime})} \arrow[rr, "\alpha_{c^{\prime}}"', bend right, color=nattran] & \textcolor{source}{c^{\prime}} \arrow[l, "F"', color=functor] \arrow[r, "G", color=functor] & \textcolor{target}{G(c^{\prime})}          
    \end{tikzcd}
\end{equation}
\note{unpack this a drop more}

\subsection{Sheaves}
\label{sec:related-work:continuity:sheaves}
Sheaves provide a mathematical abstraction of data, the space it lives over, and relationships between subsets. Sheaves can be thought of as an "algebraic data structure" \cite{ghristElementaryAppliedTopology2014} for data that lives over topological spaces. A presheaf  $\sheaf$ applied to fiber bundles is a map from an open set in a topological space, such as $\openset \subset \dbase$ where $(\mathcal{T}, \dbase)$ is a topological space, to the set of sections of a bundle $\dtotal$ over that openset \cite{spanier1989algebraic,SheafMathematics2021,nlab:presheaf}.
\begin{equation}
  \label{eq:atct:presheaf}
  \sheafc: \opensetc \textcolor{sheaf}{\rightarrow} \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}
\end{equation}
The presheaf preserves the inclusion maps $\iota$ between open sets and the restriction maps $\iota^*$ between the sets of sections over that open sets. 
\begin{equation}
  \label{eq:related-work:continuity:presheaf}
  \begin{tikzcd}[column sep=small]
    {\cgamma{\opensetc_1}{\dtotalc\restriction_{{\opensetc_1}}}} & {} & {\cgamma{\opensetc_2}{\dtotalc\restriction_{\opensetc_2}}} \arrow[ll, "\iota^*"', hook', color=set] \\& &\\
    \opensetc_1 \arrow[uu, "\sheafc_{\dbasec}", color=sheaf] \arrow[rr, "\iota"', hook, color=base] & {} \arrow[uu, "\sheafc_{\dbasec}" description, dotted, color=sheaf] & \opensetc_2 \arrow[uu, "\sheafc_{\dbasec}"', color=sheaf]                    
    \end{tikzcd}
\end{equation}

This means that a smaller space $\openset_1$ is included in a larger space $\openset_2$ and a function in a set of global sections over a bundle $\dsection \in \Gamma(\openset_2, \dtotal\restriction_{\openset2})$ that is continuous over a larger space $\openset_2$ is continuous when restricted to a subspace $\dsection \in \Gamma(\openset_1,\dtotal\restriction_{\openset2}),\openset_1 \subset\openset_2$.

A presheaf $\sheaf$ over a limit of open sets that contain a point $\dbasepoint \in \dbase$ is approximately the same as a sheaf over a point \dbasepoint\ and is called the stalk\cite{StalkSheaf2019} 
\begin{equation}
  \label{eq:atct:stalk}
  \sheafc(\dbasec)\restriction_{\dbasepointc}\coloneqq \lim\limits_{\opensetc\ni \dbasepointc} \Gamma(\opensetc, \dtotalc\restriction_{\opensetc}) 
\end{equation}

Sheaves are presheafs where sections are continuous over all unions of open sets in $\dbase$\cite{harder2008lectures, spanier1989algebraic, nlab:sheaf}. This means that a section which is defined over a stalk $\dsection \in \sheaf(\dbase)\vert_{\dbasepoint_i}$ and evaluated at a point $\dbasepoint_j$ is equal to a section in the stalk over that point $\dsection_{\dbasepoint_i}\vert_{\dbasepoint_j} = \dsection_{\dbasepoint_j}$

In category theory, a presheaf \sheafc\, is generalized to be a contravariant functor\cite{nlab:presheaf} from a category to a set. 
\begin{equation}
  \label{eq:actc:sheaf}
 \sheafc: \textcolor{source}{\mathcal{C}^{op}} \textcolor{sheaf}{\rightarrow} \setc
\end{equation}

Sheaves can be moved over different topological spaces if there is a map between those spaces, which we can use to describe how sections of a sheaf over one space behave when moved over to a different space and the maps between these spaces. Given a map $\textcolor{functor}{f}:\textcolor{source}{X} \textcolor{functor}{\rightarrow} \textcolor{target}{Y}$ between topological spaces $\textcolor{source}{X}, \textcolor{target}{Y}$ and sheaves on each space $\sheafc_{\textcolor{source}{X}}, \sheafc_{\textcolor{target}{Y}}$, we can construct a pullback functor $f^*$ and pushforward functor $f_*$ \cite{harder2008lectures} that can be composed with the sheaf functors. 
\begin{equation}
  \label{eq:atct:pullback-pushforward}
  \begin{tikzcd}[row sep=huge]
    \sheafc_{\textcolor{source}{X}}(\textcolor{source}{X}) \arrow[rr, "\textcolor{functor}{f_*}", color=functor]                            &  & (\textcolor{functor}{f_*}\sheafc_{\textcolor{source}{X}})(\textcolor{target}{Y})    \\
    \textcolor{source}{X} \arrow[rr, "\textcolor{functor}{f}", color=functor] \arrow[u, "\sheaf_{\textcolor{source}{X}}", color=sheaf] \arrow[d, dashed, color=sheaf] &  & \textcolor{target}{Y} \arrow[d, "\sheaf_{\textcolor{target}{Y}}", color=sheaf] \arrow[u, dashed, color=sheaf] \\
    (\textcolor{functor}{f^*}\sheafc_{\textcolor{target}{Y}})(\textcolor{source}{X}) \arrow[rr, "\textcolor{functor}{f^*}", color=functor]  &  & \sheafc_{\textcolor{target}{Y}}(\textcolor{target}{Y})                             
  \end{tikzcd}
\end{equation}
As shown in \autoref{eq:atct:pullback-pushforward}, the pushforward $f_*\sheaf_{X}$ modifies the sheaf $\sheaf_{X}$ such that sections $s \in \sheaf_{X}$ can be evaluated over points in $y$
\begin{equation}
  \label{eq:atct:pushforward-section}
  (\textcolor{section}{s}: \textcolor{source}{x} \textcolor{section}{\rightarrow} \setc) \textcolor{functor}{\xrightarrow{\mathmakebox[3em]{f_*}}} (\textcolor{functor}{f_*}\textcolor{section}{s}: \textcolor{target}{y} \textcolor{section}{\rightarrow} \setc)
\end{equation}
which means it decorates the section function $s$ so that it can take as input values in $Y$ and return the corresponding section in $X$ 
\begin{equation}
  \label{eq:atct:pushforward-f}
  \textcolor{section}{s}(\textcolor{source}{x}) 
  = \textcolor{section}{s}\textcolor{functor}{_*}(\textcolor{target}{y}) =\textcolor{section}{s}(\textcolor{functor}{f^{-1}}(\textcolor{target}{y}))
\end{equation}
The pullback functor $f^*$ modifies the sheaf  $\sheaf_{Y}$ such that sections $t \in \sheaf_{Y}$ can be evaluated over points in $x$
\begin{equation}
  \label{eq:atct:pullback-section}
  (\textcolor{functor}{f^*}\textcolor{section}{t}: \textcolor{source}{x} \textcolor{section}{\rightarrow} \setc) \textcolor{functor}{\xrightarrow{\mathmakebox[3em]{f^*}}} (\textcolor{section}{t}: \textcolor{target}{y} \textcolor{section}{\rightarrow} \setc)
\end{equation}
which means it decorates the section function $t$ so that it can take as input values in $X$ and return the corresponding section in $Y$ 
\begin{equation}
  \label{eq:atct:pullback-f}
  \textcolor{section}{t}\textcolor{functor}{^*}(\textcolor{source}{x})
  =\textcolor{section}{t}(\textcolor{functor}{f}(\textcolor{source}{x}))
  =\textcolor{section}{t}(\textcolor{target}{y}) 
\end{equation}
The pushforward and pullback functors are isomorphic \cite{harder2008lectures}
\begin{equation}
  \label{eq:atct:pullback-pushforward-isomorphic}
  \textcolor{homset}{hom}_{\sheafc_{\textcolor{source}{X}}}(\textcolor{functor}{f*}\sheafc_{\textcolor{target}{Y}}, \sheafc_{\textcolor{source}{X}}) = 
  \textcolor{homset}{hom}_{\sheafc_{\textcolor{target}{Y}}}(\sheafc_{\textcolor{target}{Y}}, \textcolor{functor}{f_*}\sheafc_{\textcolor{source}{X}})
\end{equation}
such that the diagram is commutative
\begin{equation}
  \label{eq:pullback-pushforward-homset}
  \begin{tikzcd}
    \sheafc_{\textcolor{source}{X}}(\textcolor{source}{X}) \arrow[rr, "\textcolor{functor}{f_*}", color=functor]  &  & (\textcolor{functor}{f_*}\sheafc_{\textcolor{source}{X}})(\textcolor{target}{Y}) \\
    && \\
    (\textcolor{functor}{f^*}\sheafc_{\textcolor{target}{Y}})(\textcolor{source}{X}) \arrow[rr, "\textcolor{functor}{f^*}", color=functor] \arrow[uu, "\textcolor{homset}{Hom_{\sheaf_X}}", color=homset] &  & \sheafc_{\textcolor{target}{Y}}(\textcolor{target}{Y}) \arrow[uu, "\textcolor{homset}{Hom_{\sheaf_Y}}"', color=homset] \arrow[lluu, "\textcolor{homset}{Hom_{XY}}"' description, dashed, color=homset]
    \end{tikzcd}
\end{equation}
meaning that transformations of sections of one sheaf to sections of another are equivalent regardless of the space the sheaf is over.

\section{Artist} 
In this section we describe the properties a data to visualization transform must satisfy to be considered equivariant and continuity preserving. We propose that by explicitly defining these properties, we can better incorporate them into visualization library design.

We formulate the visualization transformation as a functor, which we call the \textcolor{artist}{Artist} \vartist \footnote{We call this transformation the artist because the \mintinline{python}{Artist} object in Matplotlib\cite{hunterMatplotlib2DGraphics2007}}. The artist \vartist\ converts \textcolor{total}{data}, which we denote $\mathcal{\dtotalc}$ to \textcolor{total}{graphics} $\mathcal{\gtotalc}$. We formulate this association as

\begin{equation}
  \label{eq:artist:artist}
  \vartistc: \mathcal{\dtotalc} \rightarrow \mathcal{\gtotalc}
\end{equation}

In this section we encapsulate the structure of the data and graphic spaces as categories. We then use these definitions to express the structure that an artist preserves. Finally, we introduce rules for composing artists in a way where the structure between the inputs to the individual artists is preserved.

\subsection{Continuity}
\label{sec:artist:category}

Fiber bundles, as introduced in \autoref{sec:related-work:fiber-bundles}, serve as the model of data and graphic spaces in out model. In this section, we introduce categorical formulations of the topological spaces that make up a fiber bundle.

We introduce the category $\mathcal{\dbase}$ to encapsulate the continuity of the data, for example if the data are discrete points, continuous lines, images, or networks. 
\begin{definition} \label{def:category:K} The category of open sets $\mathcal{\dbasec}$ consists of
  \begin{itemize}
    \item \textit{objects} open sets $\openset_{i} \in \{\openset\}$ in the topological space $(\dbase, \mathscr{T})$), including the empty set $\varnothing$ and the maximum set \dbase. 
    \item \textit{morphisms} $\iota: \openset_{i} \hookrightarrow \openset_{j}$ for all  $\openset_i, \openset_j \subset \dtotal$
  \end{itemize} 
\end{definition}

We also introduce an analogous category $\mathcal{\gbase}$ that encapsulates the topology of the graphic, which is the topology of the rendering in the display space; for example, this rendering can be a 2D or volumetric image or a 3D print.

\begin{definition} The category of open sets $\mathcal{\gbasec}$ in the topological space $(\dbase, \mathscr{T})$ consists of open set \textit{objects} $\opensetg \in \{\opensetg\}$, including the empty set $\varnothing$ and the maximum set $\gtotal$ and \textit{morphisms} $\iota: \opensetg_i \rightarrow \opensetg_j$.  
\end{definition}

The data space \dbase\ is a deformation retract of the graphic topology \gbase\meaning that regions in \gbase\ collapse to points in \dbase\ such that the relative positions of the points are preserved \cite{nlab:deformation_retract}. One method of constructing the graphic space is by expanding the dimensionality of the data \dbase\ to match the spatial dimensionality of the rendering by multiplying the data space by  intervals $I$.

\begin{figure}[h!]
  \label{fig:artist:gbase}
  \includegraphics[width=\columnwidth]{deform_retract.png}
  \caption{Given a 2D screen display, the graphic space \gbase\ must be 2D. It is constructed by multiplying the line \dbase\ by the interval [0,1] such that every vertical band in \gbase\ corresponds to a unique point $\dbasepoint \in \dbase$.}
\end{figure}

For example, as illustrated in \autoref{fig:artist:gbase}, the display space is a 2D screen and the data space is a 1D line \dbase. For that line to have any thickness, it must be rendered as a ban. One method of obtaining this graphic continuity while preserving the constraint that it is a deformation retract is by multiplying the base space \dbase\ by an interval $I=\[0,1\]$ such that $\gbase = \dbase \times I$. This construction ensures that every region $(\dbasepoint, \alpha)\vert_{\alpha \in I} \subset \gbase$ maps to a  corresponding point $\dbasepoint \in \dbase$.

\subsubsection{Data}
\note{clean up why sheaf}
As mentioned in \autoref{sec:related-work:continuity:sheaves}, sheaves are a powerful tool for keeping track of data that lies over topological spaces. Therefore, we model the type of data an artist expects as a sheaf $\sheafc_{\dtotalc}: \mathcal{\dbasec}^{op} \textcolor{sheaf}{\rightarrow} \setc$.
\begin{equation}
  \label{eq:artist:morphism-of-sheaf-data}
  \begin{tikzcd}
    & \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \in Ob(\setc) & \\
    {} & & {} \\
    & \opensetc \in Ob(\mathcal{\dbasec}^{\textcolor{base}{op}}) \arrow[uu, "\sheafc_{\dtotalc}"', shift left = 10, color=sheaf] &   
\end{tikzcd}
\end{equation}
The objects of \textbf{Set} in this sheaf are the sets of sections $\Gamma(\dbase, \dtotal)$ introduced in \autoref{eq:related-work:continuity:sections} and the morphisms are the inclusion maps $(\iota, \iota^*)$ introduced in \autoref{eq:related-work:continuity:presheaf}.

To establish a mapping between graphic and data base spaces, we introduce the surjective map $\vindex: \gbase \rightarrow \dbase$. 
\begin{equation}
  \label{eq:artist:vindex}
  \begin{tikzcd}
    \dtotalc& \dtotalpullc \arrow[l, "\vindexpullc"', dashed, color=artist]                                                                           \\
    \dbasec \arrow[u, "\dsectionc", color=section] & \gbasec \arrow[l, "\vindexc", color=artist] \arrow[lu, "\dsectionc \circ \vindexc" description, color=section] \arrow[u, "\dsectionpullc"', dashed, color=section]
    \end{tikzcd}
\end{equation}
such that the data continuity $\dbase$ is a deformation retract of $\gbase$, as expressed in \autoref{eq:artist:gbase}.

>>>>

\IEEEraisesectionheading{\section{Introduction}\label{sec:intro}}

\IEEEPARstart{T}

\section{Related Work}

\section{Expressing Structure}
\label{sec:atct}
\note{why why why?}
We use concepts from algberaic topolology and category theory to formally express the structure visualization components are expected to preserve. Category theory is a mathematical langauge for describing how objects are specified, constrained, and composed \cite{wielsManagementEvolvingSpecifications1998}\; therefore in this work we propose implementable in code categorical constructions of visualization components. Fiber bundles, are a "unified, dimension-independent framework", as described by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, that allow us to seperatly describe the topololigical properties and fields of a data source and also the mapping between topolology and data values. Sheaves on bundles are an "algebraic data structure", as described by Ghrist \cite{ghristElementaryAppliedTopology2014}, for for expressing the bookeeping involved in subsetting data and composing data that is distributed or streaming.

\subsection{Fiber Bundles}
\label{sec:atct:fiber-bundles}

\section{Preserving Structure}
